---
title: "M4 Data Analysis: One-Sample Procedures"
author: "Matthew Ragoza"
date: "10/12/2021"
always_allow_html: true
output:
  html_document:
    df_print: paged
    toc: true
    number_sections: true
    toc_depth: '3'
    code_folding: show
  github_document:
    toc: true
---

```{r,echo=FALSE,message=FALSE,warning=FALSE}
require(knitr)
# Set so that long lines in R will be wrapped:
opts_chunk$set(tidy.opts=list(width.cutoff=100),tidy=TRUE)
library('ggplot2')
library('dplyr')
library('tidyverse')
library('kableExtra')
library('DescTools')
```

# CD4 cell counts

The purpose of this analysis is to determine if the mean CD4 cell count of HIV positive patients is less than 200 before they receive treatment. The population consists of HIV patients prior to receiving medical treatment, while the sample is a set of 29 HIV-positive patients from this population, who also have not received treatment.

The distribution of CD4 cell count in this sample can be seen in Figure 1 and has been summarized numerically in Table 1. The cell count is a quantitative variable, so it has been displayed as a box plot, and summarized using measures of center and spread. The sample mean of 188.1 is less than the sample median of 194.6, indicating a left-skewed distribution.

```{r, fig.width=5, fig.height=2}
load('cd4.RData')
ggplot(cd4 %>%
    drop_na(cd4),
    aes(x=cd4)
) + geom_boxplot() + theme_classic() + theme(axis.text.y=element_blank()) + xlab('CD4 cell count')
```

**Figure 1.** Visualization of CD4 cell count distribution.

```{r}
cd4 %>% 
summarize(
    N=n(),
    Mean=mean(cd4, na.rm=TRUE),
    SD=sd(cd4, na.rm=TRUE),
    Q1=quantile(cd4, 0.25, na.rm=TRUE),
    Median=median(cd4, na.rm=TRUE),
    Q3=quantile(cd4, 0.75, na.rm=TRUE),
    IQR=quantile(cd4, 0.75, na.rm=TRUE)-quantile(cd4, 0.25, na.rm=TRUE),
    Min=min(cd4, na.rm=TRUE),
    Max=max(cd4, na.rm=TRUE),
    Range=max(cd4, na.rm=TRUE)-min(cd4, na.rm=TRUE),
) %>%
format(digits=4) %>%
kable() %>%
kable_styling()
```
**Table 1.** Numerical summary of CD4 cell count distribution.


Based on this sample, we want to know if the population mean CD4 cell count of pre-treatment HIV patients is significantly lower than 200. Therefore, the null hypothesis is that the population mean CD4 cell count is 200, and the alternative hypothesis is that the population mean CD4 cell count is less than 200.

To test this hypothesis, we first need to decide an appropriate hypothesis test to use. We have a simple random sample, but our sample size is only 29, which is just below the rule of thumb value of 30. However, if CD4 cell counts are normally distributed in the population, then it is still acceptable to apply a t-test. We can check this by using the normal quantile plot in Figure 2 to see if CD4 cell counts in the sample are approximately normal. Since the CD4 cell count quantiles in the sample are very close to the identity line when plotted against the quantiles of a theoretical normal distribution, the sample is approximately normally distributed. Therefore, the requirements are satisfied to apply a t-test.

```{r, fig.width=4, fig.height=4}
ggplot(cd4,
    aes(sample=cd4)
) + geom_qq() + geom_qq_line() + theme_classic() + ylab('CD4 cell count quantiles') + xlab('Normal theoretical quantiles')
```

**Figure 2.** Normal quantile plot of CD4 cell count to determine normality.

Before proceeding, we will decide on a significance level of 0.05. We will now run the t-test.
```{R}
alpha = 0.05
t.test(x=cd4, mu=200, alternative='less')
```
The test statistic indicates that the observed sample mean is -1.521 standard errors from the null hypothesis value of CD4 cell count. The p-value indicates that we would have a 6.98% probability of observing a sample mean at least as extreme as this if the null hypothesis were true.

Since we previously set the significance level to 0.05, the p-value of 0.0698 is greater than the required significance level. Therefore, we fail to reject the null hypothesis, as there is insufficient evidence to support that the population mean CD4 cell count of HIV patients is less than 200 prior to treatment.

```{R}
n = nrow(cd4)
xbar = mean(cd4$cd4)
s = sd(cd4$cd4)
moe <- qt(1-alpha, df=n-1)*s/sqrt(n)
c(xbar-moe, xbar+moe)
```
The 90% confidence interval for the population mean of CD4 cell count is [174.8, 201.4]. This indicates that if we were to take repeated samples from the population and construct confidence intervals like this, 90% of the time they would capture the true population mean. In context of the study, there is a 90% probability that the CD4 cell count range from 174.8 to 201.4 contains the true population mean for pre-treatment HIV patients. This interval contains the hypothesized value of 200, so it agrees with the result of the hypothesis test.

```{R}
delta = 10
s = sd(cd4$cd4)
get_power_n <- function(power) {
    ceiling(power.t.test(power=power, delta=delta, sd=s)$n)
}
df <- data.frame(power=c(0.5,0.6,0.7,0.8,0.9))
df$n <- sapply(df$power, FUN=get_power_n)
kable(df) %>% kable_styling()
```
**Table 2.** Sample sizes required for a given power in the follow-up study.

If the researchers would like to detect a difference in CD4 cell counts of 10 in their follow-up study, Table 2 shows the sample sizes that would be required for different powers. Power is the probability of correctly rejecting a false null hypothesis, and is generally desired to be at least 80%. Therefore, the researchers should try to obtain a sample size of at least 280 participants for their follow-up study to achieve a power of 80%.

# Lung cancer rate

Next we analyze the proportion of deaths that are due to lung cancer among workers in a chemical plant. The population was men aged 15-64 in England and Wales who died between 1970-1972, and the population proportion of deaths due to lung cancer was 12%. The sample was a group of 40 men in this group who worked for at least one year in a chemical plant by the time they died. Of those 40 deaths among chemical plant workers, 10 were due to lung cancer, which is a sample proportion of 25%.

```{R}
n = 40
x = 10
p0 = 0.12
```

We would like to determine if the proportion of deaths due to lung cancer was significantly higher among former chemical plant workers than the broader population proportion. Therefore, the null hypothesis is that the proportion of lung cancer deaths among chemical plant workers was the same as the population proportion of 12%. The alternative hypothesis is that the proportion of lung cancer deaths in chemical plant workers is greater than 12%.

To determine the appropriate hypothesis test to apply, we first have to check the requirements. We have a simple random sample, however we do not have the requirement that that we would expect at least five successes in the sample under the null hypothesis (n*p0 = 4.8), as seen below. Therefore, we will use an exact binomial test instead of a z-test.
```{R}
c(n*p0, n*(1-p0))
```

Again, we will use a significance level of 0.05. We can now run the test.
```{R}
alpha = 0.05
binom.test(x, n, p0, alternative='greater')
```
Since a binomial test was run instead of a z-test, there is no test statistic to interpret. However, the p-value indicates that we have a 1.75% chance of observing a sample proportion of deaths due to lung cancer at least as extreme as this, assuming that the null hypothesis is true.

The p-value of 0.0175 is less than the significance level of 0.05, so we reject the null hypothesis. Therefore, this is evidence that the population proportion of lung cancer deaths among chemical plant workers is significantly greater than the 12% rate of the general population.

```{R}
BinomCI(x, n, conf.level=0.90, method='clopper-pearson')
```
The 90% confidence interval for the population proportion of deaths due to lung cancer is [14.2%, 38.7%]. This indicates that if we were to take repeated samples from the population and construct confidence intervals like this, 90% of the time they would capture the true population proportion. In context of the study, there is a 90% probability that the interval from 14.2% to 38.7% contains the true population proportion of deaths due to lung cancer among former chemical plant workers in this demographic. This interval does not contain the general population proportion of 12%, so it agrees with the result of the hypothesis test.