---
title: "M5 Data Analysis: Contingency Tables"
author: "Matthew Ragoza"
date: "10/19/2021"
always_allow_html: true
output:
  github_document:
    toc: true
  html_document:
    df_print: paged
    toc: true
    number_sections: true
    toc_depth: '3'
    code_folding: show
---

```{r,echo=FALSE,message=FALSE,warning=FALSE}
require(knitr)
# Set so that long lines in R will be wrapped:
opts_chunk$set(tidy.opts=list(width.cutoff=100),tidy=TRUE)
library('ggplot2')
library('dplyr')
library('tidyverse')
library('kableExtra')
library('DescTools')
library('janitor')
```

# Representation in Research

We are interested in investigating racial representation in clinical trials to determine if it matches the racial makeup of the broader population. A numerical summary of the distributions of race in the population and in a sample of lung cancer clinical trial participants is shown in Table 1. This information is also shown as side-by-side bar plots in Figure 1 to facilitate comparison of the racial breakdown of the population and the clinical trial sample.

```{R}
load('LungCancer.RData')

agg <- lungcancer %>%
    group_by(race) %>%
    summarize(count=n()) %>%
    mutate(ClinicalTrials=round(100*count/sum(count), digits=2)) %>%
    arrange(-count)

agg$Population <- c(75.6, 10.8, 9.1, 3.8, 0.7)
agg$expected <- round(agg$Population/100 * sum(agg$count), 2)

agg %>%
    adorn_totals() %>%
    kable(col.names=c(
        'Race',
        'Clinical trial count',
        'Clinical trial proportion (%)',
        'Population proportion (%)',
        'Population-representative count'
    )) %>%
    kable_styling()
```
**Table 1.** Numerical summary of race distribution in overall population and in lung cancer clinical trials.

```{r}
agg %>%
    gather('key', 'value', ClinicalTrials, Population) %>%
    ggplot(aes(
        x=factor(race, levels=agg$race), 
        y=value,
        fill=factor(key, levels=c('Population', 'ClinicalTrials'))
    )) +
    geom_bar(position='dodge', stat='identity') +
    xlab('Race') +
    ylab('Proportion (%)') +
    ylim(0, 100) +
    theme_classic() +
    theme(legend.title=element_blank())
```

**Figure 1.** Visualization of race distribution in overall population compared to lung cancer clinical trials.

Visual assessment of Figure 1 suggests that there are differences in the racial distribution of clinical trials compared to the general population. We can state this formally in order to enable a hypothesis test: the null hypothesis is that the proportions of each racial group in lung cancer clinical trials are the same as their population proportions. The alternative hypothesis is that at least one racial group has a different proportion in clinical trials compared to their population proportion.

To test this hypothesis, a chi-squared goodness-of-fit test is appropriate. Chi-squared goodness-of-fit tests can detect whether frequencies of a categorical variable with more than two categories (race) are different in a sample (clinical trial participants) compared to a theoretical distribution (the general population). The participants in the clinical trials were likely determined by simple random sampling. Furthermore, at least five clinical trial participants of each race would be expected under the null hypothesis, as seen in the last column of Table 1. Therefore, the requirements are satisfied to apply a chi-squared test, which will now be performed using a significance level of 0.01.
```{R}
alpha = 0.01
k = nrow(agg)
chisq.test(x=agg$count, p=agg$Population/100)
qchisq(df=k-1, p=1-alpha) # the critical value
```
The chi-squared test statistic provides a measure of statistical distance between the racial distribution in clinical trials compared to the general population. Since the critical value is 13.28 for the chosen significance level of 0.01, we would expect to see chi-squared test statistics greater than 13.28 only 1% of the time if the null hypothesis were true. The observed chi-squared test statistic is 524.71, which is 40 times larger than the critical value. The p-value of 2.2e-16 indicates that we would almost never observe a test statistic this large if the null hypothesis were true. We can reject the null hypothesis because the p-value is less than the significance level. In conclusion, we have very strong evidence that the racial distribution in lung cancer clinical trials is different from the racial makeup of the general population.

# Mammography

The next question that we will investigate is whether there is a relationship between mammography experience and family history of breast cancer. A contingency table summarizing the relationship between these variables numerically is shown in Table 2. The relationship is shown graphically as stacked bar plots of relative frequencies in Figure 2.

```{R}
load('mamexp.RData')

agg <- mamexp %>%
    group_by(famhist, me) %>%
    count() %>%
    spread(key=famhist, value=n) %>%
    adorn_totals(where=c('row', 'col'))

agg %>%
    kable(col.names=c(
        'Mammography experience',
        'No family history of breast cancer',
        'Family history of breast cancer',
        'Total'
    )) %>%
    kable_styling()
```
**Table 2.** Contingency table of counts summarizing relationship between mammography experience and family history of breast cancer.

```{r, message=FALSE}
mamexp %>%
    group_by(famhist, me) %>%
    summarise(count=n()) %>%
    mutate(freq=100*count/sum(count)) %>%
    ggplot(aes(
        x=famhist, 
        y=freq,
        fill=me
    )) +
    geom_bar(position='stack', stat='identity') +
    xlab('Family history of breast cancer') +
    ylab('Relative frequency (%)') +
    labs(fill='Mammography\nexperience') +
    theme_classic() +
    theme()
```

**Figure 2.** Visualization of relationship between mammography experience and family history of breast cancer.

Based on Figure 2, there appear to be differences in frequency of mammography experience between the women in the sample with a family history of breast cancer and those without. We can test this statistically by first defining the hypotheses. The null hypothesis is that mammography experience is independent of family history of breast cancer. The alternative hypothesis is that mammography experience depends on the history of breast cancer in the participant's family.

To test this hypothesis, we can apply a chi-squared test of independence. Chi-squared tests of independence can determine if there is an association between two categorical variables, and are appropriate when the variables are two measurements from the same observational units. In this study, the variables are two different answers provided by participants on a questionnaire, so they come from the same observational units.

```{R}
N = last(agg$Total)
N_no = last(agg$no)
N_yes = last(agg$yes)
agg %>%
    mutate(
        no_expected=round(Total*N_no/N, 2),
        yes_expected=round(Total*N_yes/N, 2)
    ) %>%
    select(c('me', 'no_expected', 'yes_expected', 'Total')) %>%
    kable(col.names=c(
        'Mammography experience',
        'No family history of breast cancer',
        'Family history of breast cancer',
        'Total'
    )) %>%
    kable_styling()
```

**Table 3.** Contingency table showing expected counts assuming that mammography experience and family history of breast cancer are independent.

The design of this study probably involved a random sample of observational units (participants). Furthermore, it can be seen in Table 3 that every cell in the contingency table has a least 5 counts, assuming the null hypothesis is true. Therefore, the requirements are satisfied to apply a chi-squared test of independence, which we will now perform with significance level 0.10.

```{R}
alpha = 0.10
r = length(unique(mamexp$me))
c = length(unique(mamexp$famhist))
chisq.test(x=table(mamexp$me, mamexp$famhist), correct=FALSE)
qchisq(df=(r-1)*(c-1), p=1-alpha) # the critical value
```
In this context, the chi-squared test statistic measures the disparity between the observed joint distribution and what would be expected if mammogram experience and family history of breast cancer were independent. At the significance level of 0.10, the critical value of 4.61 means that we would have a 10% probability of seeing test statistics greater than 4.61 if the variables were independent. The observed chi-squared test statistic is 13.05, which is larger than the critical value. Furthermore, the p-value of 0.0015 indicates that we would have a 0.15% chance of seeing a test statistic this large if the null hypothesis were true. Since the p-value is less than the significance level, we can reject the null hypothesis. In conclusion, we have strong evidence that mammography experience is dependent on family history of breast cancer.

```{R}
p_me_fh = (11+19)/44                    # P(mam_exp|fam_hist)
p_me_nfh = (63+85)/368                  # P(mam_exp|~fam_hist)
odds_me_fh = p_me_fh / (1 - p_me_fh)    # Odds(mam_exp|fam_hist)
odds_me_nfh = p_me_nfh / (1 - p_me_nfh) # Odds(mam_exp|~fam_hist)
odds_me_fh / odds_me_nfh                # odds ratio
```
The odds ratio of mammography experience for women with and without a family history of breast cancer is 3.19. This can be interpreted by saying that women with a family history of breast cancer have 3.19 times greater odds of having had a mammogram than women with no family history of breast cancer.

# Hip Protectors

The third analysis is related to the effectiveness of hip protectors in preventing hip fractures in the elderly. The observational units of the study were nursing home residents who each wore a hip protector on one hip, and the variables that were measured were whether they had a hip fracture on either the protected or unprotected hip.

We want to know from this data whether there was a difference in the rate of fractures between the protected and unprotected hips. Since the values of the two variables come from the same person, they are dependent. Therefore we treat them as paired observations, and focus on the discordant pairs. Specifically, the null hypothesis is that there was no difference between the proportion of individuals who had an unprotected hip fracture, but not a protected hip fracture, and the proportion who had a protected hip fracture, but not an unprotected hip fracture. The alternative hypothesis is that the proportion who only an unprotected hip fracture is different from the proportion that only had a protected hip fracture.

```{R}
hip <- matrix(c(309, 15, 10, 2), nrow=2)
hip
```

Given that we have two dependent categorical variables, we can test the effectiveness of the hip protectors by using McNemar's test. This test compares proportions between discordant pairs of values of two dependent categorical variables. Since the study was a randomized controlled trial, we have a simple random sample of participants. In addition, we have the requirement that the total count of discordant pairs is at least 10. Therefore we can now apply McNemar's test, using a significance level of 0.05.

```{R}
alpha = 0.05
mcnemar.test(hip, correct=FALSE)
```
Here the chi-squared statistic of 1.0 represents the statistical distance between the probability of only having a fracture in the unprotected hip and the probability of only having a fracture in the protected hip. The p-value of 0.317 indicates that we would have a 31.7% chance of seeing a chi-squared test statistic this drastic if the null hypothesis were true. Since the p-value is greater than the significance level of 0.05, we fail to reject the null hypothesis. There is insufficient evidence to conclude that the hip protectors were effective in preventing hip fractures in the elderly.
